---
title: "Problem Set 1 of API-209"
author: "Manuel Alcala Kovalski assisted by Dan Levy"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: cayman 
    highlight: github
    fig_width: 4
    fig_height: 6
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  results = 'asis',
  tidy.opts=list(width.cutoff=120),
  tidy=TRUE, 
  echo = FALSE,
   dev = "svglite",
      fig.ext = ".svg",
  #cache = TRUE,
  fig.align = "center",
  fig.show = "hold",
  fig.width=8, 
  fig.asp=0.9,
  out.width = "100%",
  fig.path = "figures/",
  fig.retina = 2,
  fig.showtext = TRUE,
  fig.pos = "t",
  dpi = if (knitr::is_latex_output()) 72 else 300,
  dev.args = list(png = list(type = "cairo-png")), 
  optipng = "-o1 -quiet"
)


library(knitr)
knitr::knit_hooks$set(optipng = hook_optipng)
knitr::knit_hooks$set(pngquant = hook_pngquant)
```

```{r packages}
librarian::shelf('tidyverse', 'readxl', 'kable', 'kableExtra', 'magrittr',  'glue', 'reactable', 'gt')
```

## Question 4:

### Learning about the world economy

```{r load-wdi, include = TRUE}
wdi_raw <-  read_xlsx('data/wdi.xlsx')

wdi_raw %>% 
  head(10) %>% 
   kbl() %>% 
  kable_styling()

wdi_raw %>% 
  names()
```

There are `r wdi_raw %>% nrow()` observations in the data set.
Some key variables are `r wdi_raw %>% names()`.

The type for each key variable and the extent of missigness is shown in the table below.

```{r  keytable}
wdi_raw  %>% 
  skimr::skim()
```

### Analysis data set

Once you have done this (no need to type answers to these questions, but do answer them), create an analysis data set in which:

-   You transform the population variables so that they are expressed in millions of people (for example, 158,000,000 should become 158).

-   You transform the gdp variables so that they are expressed in millions of dollars

-   You [keep only observations]{.ul} that have non-missing data for both population 2018 and gdp 2018.

This will be the data set you will you for the remainder of this problem set, so assign it to an object you can use.

```{r clean-data, include = TRUE , echo = TRUE}
wdi <-
  wdi_raw %>%
  drop_na(ends_with('2018')) %>%
  pivot_longer(
    cols = gdp_1993:pop_2018,
    names_to = c(".value", "date"),
    names_pattern = "(.*)_(.*)"
  ) %>%
  mutate(across(.cols  = starts_with(c('pop', 'gdp')),
                .fns = ~  .  / 1e6)) 


  wdi_raw %>%
  drop_na(ends_with('2018')) %>%
  pivot_longer(
    cols = gdp_1993:pop_2018,
    names_to = c(".value", "date"),
    names_pattern = "(.*)_(.*)"
  ) %>%
  mutate(across(.cols  = starts_with(c('pop', 'gdp')),
                .fns = ~  .  / 1e6)) 

```

```{r top-five-gdp}
wdi %>%
  arrange(desc(gdp)) %>% 
  head(5) %>% 
  kbl() %>% 
  kable_styling()
```

Now report the mean and the number of observations for gdp 2018 for this analysis data set.

```{r, mean2018}
wdi %>% 
  group_by(date) %>% 
  drop_na()  %>% 
  summarise(mean_gdp = mean(gdp, na.rm  = TRUE),
            observations =  n())  %>% 
  gt()
```

(2) Totals: Please calculate and report:

<!-- -->

a.  Total World GDP (expressed in trillions of 2010 dollars) and world population in 2018 (expressed in billions of people)

```{r sum}
wdi %>% 
  group_by(date) %>% 
  summarise(across(.cols = c(gdp, pop),
                   .fns = ~ sum(.x, na.rm = TRUE),
                   .names = 'total_{.col}')) %>% 
  gt() %>% 
  cols_label(date = 'Year',
             total_gdp = 'Total GDP',
             total_pop = 'Total population')  %>% 
  fmt_number(columns = c('total_gdp','total_pop'))
  
```

b.   Top 5 countries in terms of GDP in 2018 and their respective GDPs and top 5 countries in terms of population in 2018 and their respective populations. You may round your values.

```{r top-five-2018}

wdi %>% 
  filter(date == 2018) %>% 
  summarise(country, gdp)  %>% 
  arrange(desc(gdp))  %>% 
  slice(1:5) %>% 
  gt() %>% 
  tab_header(title = md('Richest countries in 2018'),
                  subtitle = md('As measured by GDP')) %>% 
  fmt_number(columns = c('gdp'))

wdi %>% 
  filter(date == 2018) %>% 
  summarise(country, pop)  %>% 
  arrange(desc(pop))  %>% 
  head(5) %>% 
  gt()  %>% 
    tab_header(title = md('Most populated countries in 2018'),
               subtitle = md('In millions')) %>% 
  fmt_number(columns = c('pop'))  %>% 
  cols_label(pop =  'Population',
             country  = 'Country')
```

  (3) **Central Tendencies:** Calculate GDP per capita (which is equal to total GDP divided over population, but beware of units) for each country in the database for 1993 and 2018.

```{r central-tendency}
central_tendencies  <-
  wdi %>% 
      filter(date ==  2018)  %>% 
      summarise(`Mean GDP per capita in 2018` = mean(gdp/pop, na.rm  = TRUE),
                `Median GDP per  capita in 2018`  = median(gdp /  pop, na.rm  =  TRUE),
                `World GDP  per capita in 2018` = sum(gdp, na.rm = TRUE) / sum(pop, na.rm = TRUE))

central_tendencies %>% 
  gt() 

central_tendencies %>% 
  pivot_longer(everything()) %>% 
  ggplot(aes(x = forcats::fct_reorder(name, value), y = value, color = name))  +
  geom_point(size = 4) +
  geom_segment(aes(xend = name, yend = 0), size = 1.2) +
  coord_flip()  +
  gghutchins::theme_hutchins() +
  theme(legend.position  = 'none') +
  gghutchins::scale_color_hutchins() +
  scale_y_continuous(breaks  =  seq(0, 15000, 2500),
                     limits = c(0,  15000)) +
  labs(title = 'Central tendencies',
       subtitle =  'GDP per capita in 2018 (Millions of USD)')
```

  Then report the following statistics.
To help you code more efficiently, feel free to write code chunks that answer several of the questions below at the same time:

```{r gdp-pc}
    wdi %<>% 
      mutate(gdp_pc = gdp / pop) 
```

```{r pop-below-mean}
population_below_gdp_mean <- 
  wdi %>% filter(date == 2018) %>% filter(gdp_pc < mean(gdp_pc, na.rm = TRUE)) %>%   summarise(`Total population   below mean GDP` = sum(pop, na.rm = TRUE)  %>% round())
```

-   The total population of all countries in 2018 with GDP per capita below the world mean calculated in part (a) is `r population_below_gdp_mean`
-   Draw a histogram

```{r histogram}
wdi %>%
  pivot_longer(gdp_pc) %>% 
  filter(date == 2018) %>% 
  ggplot(aes(x = value, fill = name)) +
  geom_histogram(binwidth = 2000) +
  scale_x_continuous(labels = scales::label_dollar()) +
  labs(x = "GDP per capita",
       y = "Number of Countries") +
  ggthemes::theme_economist() +
  scale_fill_brewer(type  = 'qual',
                    palette = 'Set2', 
                    name = '') +
  theme(legend.position = 'none') +
  labs(title = 'Distribution of GDP per capita in 2018') 
  
```

```{r low-income}
 wdi %<>%
  filter(date  == 2018) %>% 
  mutate(income_group = fct_relevel(income_group,
                                    "Low income",
                                    "Lower middle income",
                                    "Upper middle income",
                                    "High income"
  )) 

income_proportion <- function(data){
  data %>% 
    group_by(income_group) %>% 
    summarize(n = n()) %>%
  mutate(prop = n/sum(n)) %>%
  summarise(Income = income_group, 
         `Number of countries` = n,
         Proportion = glue('{ round(prop * 100, 1) }%')) 
}
wdi %>% 
  income_proportion() %>% 
  gt() %>% 
  tab_header(title = 'Proportion of countries in each income categroy')

```

b.  What percent of countries in Sub-Saharan Africa are low income?

`r wdi %>% filter(region == 'Sub-Saharan Africa') %>% income_proportion() %>% pull(Proportion) %>% head(1)` percent of countries in Sub-Saharan Africa are low income

c.  What share of low-income countries are in sub-Saharan Africa?

```{r low-income-africa}
numerator <- wdi %>% filter(region == 'Sub-Saharan Africa') %>% income_proportion() %>% 
  pull(`Number of countries`) %>% head(1)


 denominator <-  wdi %>% income_proportion() %>% 
  pull(`Number of countries`) %>% head(1)
 
share_low_income <- glue('{round(numerator/denominator, 2) * 100}%')
```

The majority of low-income countries are in sub-Saharan Africa.
In particular `r share_low_income` of low income countries are in sub-Saharan Africa.

## Problem 5:

### Dealing with missing data

```{r clean-again}
wdi <-
  wdi_raw %>% 
  drop_na(ends_with('2018')) %>% 
   pivot_longer(cols = gdp_1993:pop_2018,
                   names_to = c(".value", "date"), 
                   names_pattern = "(.*)_(.*)") %>% 
    mutate(across(.cols  = starts_with(c('pop', 'gdp')),
                .fns = ~  .  / 1e6))

```

The database does not have a 1993 GDP figure for `r wdi %>%   filter(date == 1993) %>%    naniar::n_miss()`.

Ignoring missing values, the mean GDP per capita in 1993 was `r wdi %>%    filter(date == 1993) %>%    summarise(mean(gdp / pop, na.rm = TRUE) %>% round()) %>% pull()`

```{r missing-bar-chart}
wdi_raw %>% 
      drop_na(ends_with('2018')) %>% 
  mutate(missing_1993 = if_else(is.na(gdp_1993), 'Missing', 'Not missing')) %>% 
  group_by(missing_1993) %>% 
  summarise(mean_gdp = mean(gdp_2018 / pop_2018
                            , na.rm = TRUE) ) %>% 
  pivot_longer(mean_gdp) %>% 
  ggplot(aes( x = missing_1993, y = value, fill = missing_1993)) +
  geom_col(width = 0.5) +
  scale_y_continuous(labels = scales::dollar_format(),
                     name = '') +
  scale_x_discrete(name  = '') +
  ggthemes::theme_hc() +
  theme(legend.position = 'none') +
  gghutchins::scale_fill_hutchins() +
  labs(title =  'Mean  GDP is lower for countries with missing data',
       subtitle =  'Mean GDP in 2018 for countries with missing data in 1993') +
  gghutchins::theme_hutchins() +
  theme(legend.position = 'none')

```
