---
title: "PS2"
author: "Manuel Alcalá Kovalski assisted by Dan Levy"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: cayman 
    highlight: github
    fig_width: 4
    fig_height: 6
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  results = 'asis',
  tidy.opts=list(width.cutoff=120),
  tidy=TRUE, 
  echo = FALSE,
   dev = "svglite",
      fig.ext = ".svg",
  #cache = TRUE,
  fig.align = "center",
  fig.show = "hold",
  fig.width=8, 
  fig.asp=0.9,
  out.width = "100%",
  fig.path = "figures/",
  fig.retina = 2,
  fig.showtext = TRUE,
  fig.pos = "t",
  dpi = if (knitr::is_latex_output()) 72 else 300,
  dev.args = list(png = list(type = "cairo-png")), 
  optipng = "-o1 -quiet"
)


library(knitr)
knitr::knit_hooks$set(optipng = hook_optipng)
knitr::knit_hooks$set(pngquant = hook_pngquant)
```

```{r packages}
librarian::shelf('tidyverse', 'readxl', 'kable', 'kableExtra', 'magrittr',  'glue', 'reactable', 'gt')
```

```{r data}
wdi <- haven::read_dta('data/inequality.dta')
```

## Part A
First, start with similar metrics to what you worked with in the past problem set. Calculate the mean, median, 25th percentile value, and 75th percentile value for GDP per capita (PPP-adjusted) in 2017, using the WDI data. What does the fact that the mean is higher than the median tells us about the international distribution of per capita output?

There are several ways of answering this question.

1.  Using the `stats` package we could do:

    ```{r, echo  = TRUE}
     stats::fivenum(wdi$gdppc2017) 
    ```

2.  With `dplyr` we have several options. The most straightforward but most verbose way would be to use `summarise()` and apply the functions to `gdppc2017` individually
```{r, echo = TRUE}
wdi %>% 
  summarize(Min = min(gdppc2017, na.rm = TRUE),
            Q1 = quantile(gdppc2017, .25, na.rm = TRUE),
            Avg_gdppc2017 = mean(gdppc2017, na.rm =  TRUE), 
            Q3 = quantile(gdppc2017, .75, na.rm = TRUE),
            Max = max(gdppc2017, na.rm = TRUE)
  ) %>% 
  gt()
```
We can do the same thing more efficiently  using `across()`.

```{r}
wdi %>% 
  summarize(across(.cols  = gdppc2017,
                   # Use min, quantile (twice), mean, and max
                   # Note: since we're using two different quantile values we can                     use lamda style functions with ~ to gain more control over the                     arguments.
                   .fns = list(Min = min,
                               Q1 = ~ quantile(., 0.25, na.rm = TRUE),
                               Average = mean,
                               Q3   = ~ quantile(., 0.75, na.rm = TRUE),
                               Max = max ),
                   na.rm = TRUE)) %>% 
  gt()
```

If we wanted to extend this five number summary to all the numeric variables we could use `tidyr::pivot_longer()` and `dplyr::across()` together.

```{r}
wdi %>% 
  pivot_longer(-country) %>% 
  group_by(name) %>% 
    summarize(across(.cols  = value,
                   # Use min, quantile (twice), mean, and max
                   # Note: since we're using two different quantile values we can                     use lamda style functions with ~ to gain more control over the                     arguments.
                   .fns = list(Min = min,
                               Q1 = ~ quantile(., 0.25, na.rm = TRUE),
                               Average = mean,
                               Q3   = ~ quantile(., 0.75, na.rm = TRUE),
                               Max = max ),
                   na.rm = TRUE,
              .names  = '{.fn}')) %>% 
  kbl(digits=1) %>% 
  kable_styling()
```


(b) Inequality measures are a useful way to capture and compare information about the nature of a distribution in a single value. The inequality measure you are likely most familiar with is the Gini coefficient, which is commonly used to measure within-country inequality. The Theil index is one that can be used to measure inequality across countries. Using the dineq R package, calculate the Theil index for 1992 and 2017, based on PPP-adjusted GDP per capita. Has "international inequality" risen or fallen in the past 25 years?

```{r}
librarian::shelf('dineq')

wdi %>% 
  summarise(across(starts_with('gdppc'),
                ~ dineq::theil.wtd(.x))) %>% 
  kbl(digits = 2) %>% 
  kable_styling()

```

Therefore, international inequality has fallen over the  last 25 years.

(c) Now, let's add weights to the countries in our dataset. Calculate the population-weighted Theil index for 1992 and 2017 based on PPP-adjusted GDP per capita. How do your results change compared to the unweighted values, and why?

```{r weights}
variables <- c("pop1992",  "pop2017", "gdppc1992", "gdppc2017")
wdi %>% 
  pivot_longer(cols = all_of(variables),
               names_to =  c('.value', 'year'),
               names_pattern = "(.*)([0-9]{4})") %>% 
  group_by(year) %>% 
  summarise(gdppc_weighted_theil = dineq::theil.wtd(gdppc, pop)) %>% 
  kbl(digits = 2) %>% 
  kable_styling()
```

(d) Now explore the Gini variable in the dataset. List the highest countries with the 3 highest and 3 lowest Gini scores. Also, find the mean and median of these values, and note any broad patterns you see browsing through.

The countries with the three highest Gini coefficients are `r wdi %>% summarise(country, gini) %>% arrange(desc(gini)) %>% head(3) %>% pull(country)` while the countries with the three lowest Gini coefficients are `r wdi %>% summarise(country, gini) %>%  drop_na(gini) %>%  labelled::remove_labels() %>% arrange(desc(gini)) %>% tail(3) %>% pull(country)`

```{r gini}

wdi %>% 
  summarise(across(.cols  = gini,
                   .fns = list(Mean = mean, Median = median),
                   na.rm = TRUE,
                   .names = "{.fn}_{.col}")) %>% 
  kbl(digits = 1) %>% 
  kable_styling()

```

```{r giniplot}
wdi %>% 
  ggplot(aes(x = gdppc2017, y = gini)) +
  geom_point() +
  geom_smooth(method  =  'loess') +
  ggthemes::theme_hc()
```

(f) Define as "corrupt" as a country that is in the group of the 20% most corrupt countries, and as "highly unequal" a country that has a Gini coefficient greater than 40. Using those binary classifications in a cross-tab, calculate P(corrupt \| highly unequal) and P(highly unequal \| corrupt) and explain the difference between these two conditional probabilities to someone intelligent but not well versed in statistics.

First, define the thresholds.

```{r binary, echo = TRUE}
binary_wdi <- 
  wdi %>% 
  mutate(corrupt = if_else(corruption > quantile(corruption, 0.2, na.rm = TRUE),
                                                 'Corrupt',
                                                 'Not corrupt'),
         unequal = if_else(gini > 40,  
                           'Highly unequal',
                           'Not highly unequal'),
         country,
         .keep = 'used')
```

We can get the contingencies in two steps:
```{r, echo = TRUE}
binary_wdi %>% 
  drop_na(unequal, corrupt) %>% 
  filter(corrupt == 'Corrupt') %>% 
  count(unequal) %>% 
  mutate(prop = n / sum(n, na.rm  = TRUE)) %>% 
  gt()

```

```{r, echo = TRUE}
binary_wdi %>% 
  drop_na(unequal, corrupt) %>% 
  filter(unequal == 'Highly unequal') %>% 
  count(corrupt) %>% 
  mutate(prop = n / sum(n, na.rm  = TRUE))  %>% 
  gt()
```

Or  we could  calculate all the contingencies in one step
```{r, echo = TRUE}
binary_wdi %>% 
  drop_na(unequal, corrupt) %>% 
  group_by(corrupt, unequal) %>% 
  count() %>% 
  group_by(corrupt) %>% 
  mutate(unequal_given_corrupt = n / sum(n, na.rm = TRUE)) %>% 
  group_by(unequal) %>% 
  mutate(corrupt_given_unequal = n / sum(n, na.rm = TRUE)) %>% 
  gt()
```


```{r}
wdi %>% 
  summarise(mean_5b = mean(gdppc1992, na.rm= TRUE))
``

